---
title: "Tutorial 1: Using the `ondisc_matrix` class"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 1: `Using the ondisc_matrix` class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This tutorial demonstrates how to use `ondisc_matrix`, the core class implemented by `ondisc`. An `ondisc_matrix` is an R object that represents an expression matrix stored on-disk rather than in-memory. We cover the topics of initialization, querying basic information, subsetting, and pulling submatrices into memory. We begin by loading the `ondisc` package.

```{r setup}
library(ondisc)
```

# Initialization

`ondisc` ships with several example datasets, stored in the "extdata" subdirectory of the package.

```{r}
raw_data_dir <- system.file("extdata", package = "ondisc")
list.files(raw_data_dir)
```

The files "gene_expression.mtx", "cell_barcodes.tsv," and "genes.tsv" together define a gene-by-cell expression matrix. We save the full paths to these files in the variables `mtx_fp`, `barcodes_fp`, and `features_fp`.

```{r}
mtx_fp <- paste0(raw_data_dir, "/gene_expression.mtx")
barcodes_fp <- paste0(raw_data_dir, "/cell_barcodes.tsv")
features_fp <- paste0(raw_data_dir, "/genes.tsv")
```

An `ondisc_matrix` consists of two parts: an HDF5 (i.e., .h5) file that stores the expression data on-disk in a novel format, and an in-memory object that allows us to interact with the expression data from within R. The easiest way to initialize an `ondisc_matrix` is by calling the function `create_ondisc_matrix_from_mtx`. We pass to this function (i) a file path to the .mtx file storing the expression data, (ii) a file path to the .tsv file storing the cell barcodes, and (iii) a file path to the .tsv file storing the feature IDs and human-readable feature names. We optionally can specify the directory in which to store the initialized .h5 file, which in this tutorial we will take to be the temporary directory.

```{r}
temp_dir <- tempdir()
exp_mat_list <- create_ondisc_matrix_from_mtx(mtx_fp = mtx_fp,
                                              barcodes_fp = barcodes_fp,
                                              features_fp = features_fp,
                                              on_disk_dir = temp_dir)
```

By default, `create_ondisc_matrix_from_mtx` returns a list of three elements: (i) an `ondisc_matrix` representing the expression data, (ii) a cell-wise covariate matrix, and (iii) a feature-wise covariate matrix. The exact cell-wise and feature-wise covariate matrices that are computed depend on the inputs to `create_ondisc_matrix_from_mtx` (see documentation for full details). The advantage of computing the cell-wise and feature-wise covariates at initialization is that it obviates the need to load the entire dataset into memory a second time.

```{r}
expression_mat <- exp_mat_list$ondisc_matrix
head(expression_mat)
cell_covariates <- exp_mat_list$cell_covariates
head(cell_covariates)
feature_covariates <- exp_mat_list$feature_covariates
head(feature_covariates)
```

The initialized HDF5 file is named `ondisc_matrix_1.h5` and is located in the temporary directory.

```{r}
"ondisc_matrix_1.h5" %in% list.files(temp_dir)
```

An important attribute of `create_ondisc_matrix_from_mtx` is that it does *not* assume that entire expression matrix fits into memory. The optional argument `n_lines_per_chunk` can be used to specify the number of lines to read from the .mtx file at a time. Additionally, `create_ondisc_matrix_from_mtx` is fast: the novel algorithm that underlies this function is highly efficient and implemented in C++ for maximum speed. Typically, `create_ondisc_matrix_from_mtx` takes aboout 5-8 minutes/GB to run. Finally, for a given dataset, `create_ondisc_matrix_from_mtx` only needs to be run once, even after closing and opening new R sessions.

# Querying basic information

We can use the functions `get_feature_ids`, `get_feature_names`, and `get_cell_barcodes` to obtain the feature IDs, feature names (if applicable), and cell barcodes, respectively, of an `ondisc_matrix`.

```{r}
feature_ids <- get_feature_ids(expression_mat)
feature_names <- get_feature_names(expression_mat)
cell_barcodes <- get_cell_barcodes(expression_mat)

head(feature_ids)
head(feature_names)
head(cell_barcodes)
```

Additionally, we can use `dim`, `nrow`, and `ncol` to obtain the dimension, number of rows (i.e., number of features), and number of columns (i.e., number of cells) of an `ondisc_matrix`.

```{r}
dim(expression_mat)
nrow(expression_mat)
ncol(expression_mat)
```

# Subsetting

We can subset an `ondisc_matrix` to obtain a new `ondisc_matrix` that is a submatrix of the original. To subset an `ondisc_matrix`, we apply the `[` operator and pass a numeric, logical, or character vector indicating the cells or features to keep. Character vectors are assumed to refer to feature IDs (for rows) and cell barcodes (for columns).

```{r}
# numeric vector examples
# keep genes 100-110
x <- expression_mat[100:110,]
# keep all cells except 10 and 20
x <- expression_mat[,-c(10,20)]
# keep genes 50-100 and 200-250 and cells 300-500
x <- expression_mat[c(50:100, 200:250), 300:500]

# character vector examples
# keep genes ENSG00000107581, ENSG00000286857, and ENSG00000266371
x <- expression_mat[c("ENSG00000107581", "ENSG00000286857", "ENSG00000266371"),]
# keep cells CGTTGGGCATGGCTGC-1 and GTAACCAGTACAGTTC-1 
x <- expression_mat[,c("CGTTGGGCATGGCTGC-1", "GTAACCAGTACAGTTC-1")]

# logical vector example
# keep all genes except ENSG00000237832 and ENSG00000229637
x <- expression_mat[!(get_feature_ids(expression_mat) 
                 %in% c("ENSG00000237832", "ENSG00000229637")),]
```

Subsetting an `ondisc_matrix` leaves the original object unchanged.

```{r}
expression_mat
```

This important property, called *object persistence*, makes programming with `ondisc_matrices` easy and intuitive. The underlying HDF5 file is not copied upon subset; instead, information is shared across `ondisc_matrix` objects, making subsets fast.

# Pulling a submatrix into memory

We can pull a submatrix of an `ondisc_matrix` into memory, allowing us to perform computations on a subset of the data. To pull a submatrix into memory, we use the `[[` operator, passing a numeric, character, or logical vector indicating the cells or features that we would like to access. The data structure that underlies an `ondisc_matrix` enables fast access to both rows and columns of the matrix.

```{r}
# numeric vector examples
# pull gene 6
m <- expression_mat[[6,]]
# pull cells 200 - 250
m <- expression_mat[[,200:250]]
# pull genes 50 - 100 and cells 200 - 250
m <- expression_mat[[50:100, 200:250]]

# character vector examples
# pull genes ENSG00000107581 and ENSG00000286857
m <- expression_mat[[c("ENSG00000107581", "ENSG00000286857"),]]
# pull cells CGTTGGGCATGGCTGC-1 and GTAACCAGTACAGTTC-1
m <- expression_mat[[,c("CGTTGGGCATGGCTGC-1", "GTAACCAGTACAGTTC-1")]]

# logical vector examples
# subset the matrix, keeping genes ENSG00000107581, ENSG00000286857, and ENSG00000266371
x <- expression_mat[c("ENSG00000107581", "ENSG00000286857", "ENSG00000266371"),]
# pull all genes except ENSG00000107581
m <- x[[get_feature_ids(x) != "ENSG00000107581",]]
```

The last example demonstrates that we can pull a submatrix of an `ondisc_matrix` into memory after having subset the matrix.

We can remember the difference between `[` and `[[` by recalling R lists: `[` is used to subset a list, while `[[` is used to access elements stored within a list. Similarly, `[` is used to subset an `ondisc_matrix`, while `[[` is used to access a submatrix stored within an `ondisc_matrix`.

# Saving and loading an `ondisc_matrix`

As discussed previously, there are two components to an `ondisc_matrix`: the HDF5 file stored on-disk, and the R object stored in memory. The latter contains a file path to the former, allowing us to interact with large expression data from within R.

To save an `ondisc_matrix`, we simply call `saveRDS` on the `ondisc_matrix` R object to create an .rds file.

```{r}
saveRDS(object = expression_mat, file = paste0(temp_dir, "/expression_matrix.rds"))
rm(expression_mat)
```

We then can load the `ondisc_matrix` by calling `readRDS` on the .rds file.

```{r}
expression_mat <- readRDS(paste0(temp_dir, "/expression_matrix.rds"))
```

We also can use the constructor of the `ondisc_matrix` class to create an `ondisc_matrix` from an already-initialized HDF5 file.

```{r}
h5_file <- paste0(temp_dir, "/ondisc_matrix_1.h5")
expression_mat <- ondisc_matrix(h5_file)
```
